---
title: "Red Wine Insights"
author: "Jason DeCambre"
date: "3/29/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('formatR')
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Red Wine Insights

### Here, we'll take a look at some of the chemical features that make these wines more or less enjoyable.

### Data from a 2009 Vinho Verde dataset from Portugal:

##### P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
##### Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.

##### Available at https://archive.ics.uci.edu/ml/datasets/wine+quality

##### Inspiration also from https://towardsdatascience.com/a-data-scientists-guide-to-buying-wine-b584c77b9e60

## Loading data and taking a look:
### For some reason they separated these with a semicolon, so we'll have to override the default comma separator

```{r}
data <- read.csv('winequality-red.csv', sep=';')
head(data)
```

## Analyzing data that we don't understand isn't the best idea, so let's establish what these mean:
-  __Fixed acidity:__ this gives the wine its structure and 'zing'...too much, and it's sour, too little, and it tastes flat and flabby
- __Volatile acidity:__ while this comes from acid, it's something you sooner smell than taste. Volatile acids become gases near room temperature and so we wouldn't expect them to add much to enjoyment
- __Citric acid:__ adds to overall acidity, usually used as an additive to improve taste and make the wine more robust to other flavours
- __Residual sugar:__ wine is alcoholic because sugar ferments into alcohol. Any sugar left behind in this process is residual, and thus gives the wine a sweet taste
- __Chlorides:__ these are salts. This often has a lot to do with the kind of soil at the vineyard
- __Sulfur:__ a little reading online says that this is often added to help prevent bacterial growth, though I don't know much about this part
- __Quality:__ subjective rating by tasters between 1 and 10

## Now it would be good to check out some summary statistics and distributions for these attributes.

```{r}
summary(data)
```

## Thankfully no NA values.

### Going on to look at the structure of the data:

```{r}
str(data)
```

### This all looks clean and well standardized. Time to look at distributions and outliers.

```{r}
hist(data$fixed.acidity)
hist(data$volatile.acidity)
hist(data$citric.acid)
hist(data$citric.acid)
hist(data$residual.sugar)
hist(data$chlorides)
hist(data$free.sulfur.dioxide)
hist(data$total.sulfur.dioxide)
hist(data$density)
hist(data$pH)
hist(data$sulphates)
hist(data$alcohol)
hist(data$quality)
```


### Most of these are positively skewed, and the "quality" ratings show the tasters generally thought most of these wines to be mediocre.

### Looking for outliers:

```{r}
boxplot(data$fixed.acidity)
boxplot(data$volatile.acidity)
boxplot(data$citric.acid)
boxplot(data$citric.acid)
boxplot(data$residual.sugar)
boxplot(data$chlorides)
boxplot(data$free.sulfur.dioxide)
boxplot(data$total.sulfur.dioxide)
boxplot(data$density)
boxplot(data$pH)
boxplot(data$sulphates)
boxplot(data$alcohol)
boxplot(data$quality)
```

### Interesting. The positive skew is coming out here, so we have lots of outliers. Since this dataset has so many, I won't do anything about the outliers just yet. Maybe what's happening at the extremes has the coolest insights.

## Heatmap to show correlations

### Now that we have an idea of how our dataset looks, we can check out the correlations between the different attributes and see what jumps out:

```{r}
library(ggplot2)
library(plotly)
library(ggcorrplot)

# Correlation matrix
corr <- round(cor(data),1)

# p-value significance of correlation matrix
ps <- cor_pmat(data)

plot <- ggcorrplot(corr,hc.order = TRUE, type = 'lower', outline.col = 'white', p.mat = ps)
ggplotly(plot)


```

### Chemistry says some of these correlations make a lot of sense:
- pH and acidity are strongly negatively correlated, since lower pH values mean greater acidity
- citric acid increases fixed acidity, so they move together
- people like alcohol, so that's what was most strongly correlated with quality :)
- alcohol and fixed (liquid) acids are less dense than water and volatile acids, so more of the former mean less of the latter

## What makes an amazing red wine?

### If we take the superb wines (8+ on the scale), the mediocre wines (4-7) and the dreadful ones (3-), we might see some differences that separate the good, the bad and the ugly.

```{r}
data$category <- ifelse(data$quality >= 8, 'superb',ifelse(data$quality <= 3, 'dreadful','mediocre'))
ggplotly(ggplot(data, aes(x=category,fill=category)) + geom_bar())
```

### Nearly all the wines are mediocre. Oh well.

## Alcohol, fixed acid and quality

```{r}
ggplotly(ggplot(data,aes(x=fixed.acidity,y=alcohol,fill=category)) + geom_point(size=data$quality))
```
### Moderate to low fixed acidity and a high alcohol content seem to make the preferred wines.

## Sweet and salt

### We don't typically think of sweet and salt for wine, but let's see if there's any useful takeway from that.

```{r}
ggplotly(ggplot(data[data$category == 'dreadful',],aes(x=residual.sugar,y=chlorides)) + geom_point(color='brown4') + geom_smooth(method=lm,color='red'))
```


```{r}
ggplotly(ggplot(data[data$category == 'mediocre',],aes(x=residual.sugar,y=chlorides)) + geom_point(color='seagreen4') + geom_smooth(method=lm,color='springgreen3'))
```

```{r}
ggplotly(ggplot(data[data$category == 'superb',],aes(x=residual.sugar,y=chlorides)) + geom_point(color='steelblue3') + geom_smooth(method=lm,color='blue4'))
```


### Now the scales of the axes and the sample sizes have a lot to do with hat we see here. But the takeaway is: if you're going to make a salty red wine, at least make it sweet enough and strong enough to cover the taste.


## Conclusion: people seem to like wines with lots of alcohol, moderate acidity and not a lot of salt -- or maybe with just enough sugar to cover it over.

